Closing the gap between statistical and scientific workflows for improved forecasts in ecology

### 2025-04-05 forecastflows.tex
continue, small modifications

\noindent \textbf{Introduction}
% Still need some work...

% Ecology has become super challenged to predict stuff for decision making
Nature is increasingly threatened by multiple, mainly human-driven, drivers of change \citep{Diaz2019}. The ongoing biodiversity loss is expected to increase in the next decades because of climate change, and will continue to alter ecosystem services and human well-being \citep{IPBES2019}. To support implementation of sustainable policies, it is critical to understand trends to date and be able to forecast future dynamics. 
% $\rightarrow$ Example of stuff to predict: population size and geographic distribution (terrestrial, freshwater, marine), extinction risk,...

% General way to do this so far... (bifurcated?)
In an era of rapidly expanding big data, robust models are essential both to estimate biodiversity trends and to forecast future changes.
Estimating global biodiversity trends depends on large-scale and long-term datasets \citep[e.g.][]{Dornelas2018}. These data, gathered opportunistically from multiple sources, are often unbalanced with massive geographic, temporal and taxonomic biases. Addressing these biases requires the use of appropriate statistical inference.
Forecasting future changes---under different plausible scenarios---generally relies on either correlative models or process-based models \citep{IPBES2019}. The latter, which focus on a mechanistic representation of ecosystem functioning, are often promoted as the most realistic approach \citep{Urban2016, Pilowsky2022}.

@@

\vspace{0.5cm}
\noindent \textbf{Scientific method and workflows}

@@

% More details on the workflow:  walk through the different steps
The first step of the workflow is to define an explicit research question and formulate some hypotheses (Fig 2). This involves making assumptions about the most influential drivers, within the specific context of our study. This step help us to think about the mechanisms that could generate the data we observe, including the observational error. Naturally, this leads to the development of a model---an ensemble of mathematical equations that encapsulates our knowledge and designed to answer our research question. The general idea is to start with a relatively simple model, that we could refine later. At this stage, prioritizing biologically meaningful parameters is crucial, as it allows us to have a sense of plausible parameter values. 
The next step is the first data simulation step. The idea is to craft some fake data that reflect our assumptions about the data structure and biases, by fixing parameter values to some values (which is straightforward if the parameters are interpretable). We then fit our model to this simulated dataset and evaluate its ability to recover the prescribed values. 
Once we are confident about our model structure, we can incorporate real data. This way, we obtain parameter estimates constrained by observations. Here, difficulty in fitting the model might indicate an inherent need for more data to address our initial question. This could lead us to either simplify our research question or---ideally---launch new data collection efforts. This lead to the second data simulation step, this time using our fitted model to generate predictions. This retrodictive check allows model output to be compared to observations. 
% adding something about parameters
Within such a workflow, forecasting emerges as a natural outcome: rather than being a final goal, it only involves jointly modeling new circumstances along with the original data.

@@

\vspace{0.5cm}
\noindent \textbf{The workflow in practice}

% How to address current issues (two case studies)
% Need an introduction here
Across the different fields of ecology---for both parameter estimation and forecasting---a systematic application of a coherent workflow holds the promise to highlight the opportunities to best reduce uncertainties through new scientific insights, toward the most critical steps. This will help refocus the debate on designing new hypothesis, formulating new questions---and guiding efforts to collect new data.

@@

% Forecasting - outline current problem
Ecological forecasting is a broad field with a diverse range of methods. Process-based modeling is often considered the gold standard in ecology \citep{Urban2016, Pilowsky2022} and beyond. Process-based models are built on explicit mathematical equations to describe (supposedly causal) relationships between environmental drivers and ecological responses. They also often incorporate empirical relationships, particularly when knowledge is incomplete or when some processes are intentionally omitted. Processes are often represented at different nested spatiotemporal scales, depending on the underlying assumptions. Model development is the central step, typically requiring several years, yet it often remains opaque from an external perspective. The step of designing the model---translating knowledge and hypotheses into mathematical equations and parameters---is often blurred with the step of model calibration (or tuning), where parameter values are inferred. Models are often treated as an accumulation of multiple submodels, each governing one or several ecological processes. Rather than being fitted as a whole, submodels are calibrated separately against specific subsets of data, and some parameters are simply prescribed (i.e. fix to a value found in the literature) or tuned to reproduce some observations or theory. The way models are currently calibrated is not a coincidence, % is coincidence the right word? 
but rather an inappropriate way to accommodate their complexity, where many parameters compensate for one another.

@@

Beyond improving the model building, the workflow also has the potential to shift how process-based models are perceived, particularly by those unfamiliar with them. The workflow could refocus attention on the research question, highlighting the ecological hypotheses that justify the use and design of the model. It would thus define a clear and limited context in which the model should apply, without always arguing about the necessity of adding more and more complexity.  % or something like : "rather than driving a constant push for increasing complexity" --- my idea is that people are always like : "oww, your model is nice, but you should add [whatever they are working on]" (classical example I heard few times: mycorrhiza! disease! pathogens!)... or "rather than constantly expanding models by adding ever more processes"? 
Process-based model would once again be a way to answer a research question---whereas today, model simulations have increasingly become a subject of study on their own.
An universal workflow offers an opportunity to bridge statistical and process-based frameworks, integrating mechanistic knowledge and leveraging robust statistical approaches \citep[e.g.][]{rounce2020quantifying}.
Ideally, applying the workflow would help to move away from the traditional process-based model paradigm, where parameters are typically assigned fixed values without properly accounting for their uncertainty. Instead, it would guide a step-by-step model fitting, parameter identification, and uncertainty quantification. Process-based models would no longer be perceived as deterministic black boxes by other researchers but rather as robust statistical frameworks encapsulating both data structure and mechanistic knowledge. This shift would present a significant challenge---as it would likely reveal many issues related to model degeneracies and data limitations before achieving robust inference. However, it would prevent modelers from making biased inferences and unfounded assumptions beyond what the data can support. And it would also be an opportunity to spread the incorporation of mechanistic assumptions beyond the process-based modeling community.

% Step back
%\item we need more data, and better question (relate this to both previous study cases)
%\item where can we best reduce uncertainties through new scientific insights?
%\item machine learning! If we change nothing, what's the point of not doing ML? ML $>$ process-based without question, and ML $>$ trends without mechanisms! Where theory is lacking, or where we are far from mechanistic understanding, you might as well do ML!
%\item (ML has a way to collect and interpret large datasets...)

In a world where machine learning is rapidly advancing, there is no point of sticking to traditional methods if no changes are made. Machine learning may surpass process-based models if the latter lack a robust estimation of their parameters and fall in a complexity trap, at the cost of their interpretability. Similarly, trend analysis, when the focus is on methodological controversies (due to the lack of an iterative workflow) rather than on a robust mechanistic foundation, offers no clear advantage over machine learning.
% If we do not take care to apply this workflow, machine learning might be the more pragmatic approach...

@@

\vspace{0.5cm}
\noindent \textbf{Wrap up: how to make it happen?}

@@

% This challenge extends beyond the scope of this paper, yet we believe that applying an intelligible workflow can promote better research ethics.
Today model development in ecology is rarely transparent, which limits how easily the research community can understand modes, and thus identify potential issues. Instead of broad inclusive conversations about how to improve models to advance our ecological understanding, a significant portion of scientific debate today has become lost in methodological considerations, but we believe our workflow provides an tractable step to fixing this. By focusing on model development more tightly tied to ecological expertise, we ague this workflow should broader the community that contributes to model development. As ecologists are increasingly expanding their computational toolkits, many field, and lab and other forms of `empirical' ecologists have the basic tools to follow this workflow to build models that better represent their ecological, and---most importantly---to interrogate them.

Advancing ecology to where most researchers see models built more flexibly from their understanding of their ecological systems will not happen rapidly without a major shift in training, however. Much of ecology still divides the world into training for those focused on several groups. First, those who will conduct field and lab studies often learn a suite of particular tests matched to particular experiment designs or to simply match their predictor and response data to a flow diagram of variable types (add EXAMPLE and reference, something like: categorical $y$ and binary $y$ means ... I dunno, chi square?) and are left adrift later when asked to simulate any sort of simple model because they were never taught what a model fundamentally is. Instead, they are expected to collaborate with others when they need more complex models, which is a second group of ecological researchers---those who focus on complex statistical models and were thus trained more in model development, but often for highly specific applications (e.g., wildlife population estimates) where they may rarely build totally new generative models. Separate from these groups, process-based modelers 
% ---often with a physical or ecophysiological background---mostly focus on simulating some mechanisms 
% hmm some people talk about simulation models (for process-based models), I agree but I'm afraid in would not be clear with the data simulation we try to highlight
%do their own thing. 
Few of these groups have integrated data simulation into their statistical or scientific workflows, which is generally reserved as a form of training needed mainly by those specializing in theoretical ecology, who often solve analytical equations but rarely link to empirical data. While specialization is valuable, we argue the fundamental training in ecology has overly-siloed these groups and prevented more rapid progress.

>>>>

\section{Introduction}

% Ecology has become super challenged to predict stuff for decision making
Nature is increasingly threatened by multiple, mainly anthropogenic, drivers of change \citep{Diaz2019}. The ongoing biodiversity loss is expected to increase in the next decades because of climate change, and will continue to alter ecosystem services and human well-being \citep{IPBES2019}. To support implementation of sustainable policies, it is critical to understand trends to date and be able to forecast future dynamics. 
% $\rightarrow$ Example of stuff to predict: population size and geographic distribution (terrestrial, freshwater, marine), extinction risk,...

% General way to do this so far... (bifurcated?)
% the topic sentence is really a repetition of previous sentence, that's baaad
In an era of rapidly expanding big data, robust models are essential for trend estimation and forecasting.
Estimating global biodiversity trends depends on large-scale and long-term datasets \citep[e.g.][]{Dornelas2018}. These data, gathered opportunistically from multiple sources, are often unbalanced with massive geographic, temporal and taxonomic biases. Addressing these biases requires the use of appropriate statistical inference.
Forecasting future changes---under different plausible scenarios---generally relies on either correlative models or process-based models \citep{IPBES2019}. The latter, which focus on a mechanistic representation of ecosystem functioning, are often promoted as the most realistic approach \citep{Urban2016, Pilowsky2022}.

@@

\section{Scientific method and workflows}

@@

% More details on the workflow:  walk through the different steps
The first step of the workflow is to define an explicit research question and formulate some hypotheses (Fig 2). This involves making assumptions about the most influential drivers, within the specific context of our study. This step help us to think about the mechanisms that could generate the data we observe, including the observational error. Naturally, this leads to the development of a model---an ensemble of mathematical equations that encapsulates our knowledge and designed to answer our research question. The general idea is to start with a relatively simple model, that we could refine later. At this stage, prioritizing biologically meaningful parameters is crucial, as it allows us to have a sense of plausible parameter values. 
The next step is the first data simulation step. The idea is to craft some fake data that reflect our assumptions about the data structure and biases, by fixing parameters to some values (which is straightforward if the parameters are interpretable). We then fit our model to this simulated dataset and evaluate its ability to recover the prescribed values. 
Once we are confident about our model structure, we can incorporate real data. This way, we obtain parameter estimates constrained by observations. Here, difficulty in fitting the model might indicate an inherent need for more data to address our initial question. This could lead us to either simplify our research question or---ideally---launch new data collection efforts. This lead to the second data simulation step, this time using our fitted model to generate predictions. This retrodictive check allows model output to be compared to observations. 
% adding something about parameters
Within such a workflow, forecasting emerges as a natural outcome: rather than being a final goal, it only involves jointly modeling new circumstances along with the original data.

@@

\section{The workflow in practice}

% How to address current issues (two case studies)
% Need an introduction here
Across the different fields of ecology---for both parameter estimation and forecasting---a systematic application of a coherent workflow holds the promise to highlight the opportunities to best reduce uncertainties through new scientific insights, toward the most critical steps. This will help refocus the debate on designing new hypothesis, formulating new questions---and guiding efforts to collect new data. Here, we illustrate how such a workflow could lead to significant improvements in two case studies: (i) estimating global biodiversity trends and (ii) forecasting future species and ecosystem dynamics using process-based models.

\subsection{Trends}
% emw03april: could shorten the first 3 paragraphs, move some text to a box?

@@

\subsection{Forecasting}

% Forecasting - outline current problem
% emw03april: move also this to a box?
Ecological forecasting is a broad field with a diverse range of methods. Process-based modeling is often considered the gold standard in ecology \citep{Urban2016, Pilowsky2022} and beyond. Process-based models are built on explicit mathematical equations to describe (supposedly causal) relationships between environmental drivers and ecological responses. They also often incorporate empirical relationships, particularly when knowledge is incomplete or when some processes are intentionally omitted. Processes are often represented at different nested spatiotemporal scales, depending on the underlying assumptions. Model development is the central step, typically requiring several years, yet it often remains opaque from an external perspective. The step of designing the model---translating knowledge and hypotheses into mathematical equations and parameters---is often blurred with the step of model calibration (or tuning), where parameter values are inferred. Models are often treated as an accumulation of multiple submodels, each governing one or several ecological processes. Rather than being fitted as a whole, submodels are calibrated separately against specific subsets of data, and some parameters are simply prescribed (i.e. fix to a value found in the literature) or tuned to reproduce some observations or theory. The way models are currently calibrated is not a coincidence, % is coincidence the right word? 
but rather an inappropriate way to accommodate their complexity, where many parameters compensate for one another.

@@

Beyond improving the model building, the workflow also has the potential to shift how process-based models are perceived, particularly by those unfamiliar with them. The workflow could refocus attention on the research question, highlighting the ecological hypotheses that justify the use and design of the model. It would thus define a clear and limited context in which the model should apply, without always arguing about the necessity of adding more and more complexity.  % or something like : "rather than driving a constant push for increasing complexity" --- my idea is that people are always like : "oww, your model is nice, but you should add [whatever they are working on]" (classical example I heard few times: mycorrhiza! disease! pathogens!)... or "rather than constantly expanding models by adding ever more processes"? 
Process-based model would once again be a way to answer a research question---whereas today, model simulations have increasingly become a subject of study on their own.
Ideally, applying the workflow would help to move away from the traditional process-based model paradigm, where parameters are typically assigned fixed values without properly accounting for their uncertainty. Instead, it would guide a step-by-step model fitting, parameter estimation, and uncertainty quantification. This shift would present a significant challenge---as it would likely reveal many issues related to model degeneracies and data limitations before achieving robust inference. But ultimately, it would prevent modelers from making biased inferences and unfounded assumptions beyond what the data can support.

@@

\section{Bareers and opportunities}

@@

% This challenge extends beyond the scope of this paper, yet we believe that applying an intelligible workflow can promote better research ethics.
% emw03april: move this somewhere else?
Today model development in ecology is rarely transparent, which limits how easily the research community can understand modes, and thus identify potential issues. Instead of broad inclusive conversations about how to improve models to advance our ecological understanding, a significant portion of scientific debate today has become lost in methodological considerations, but we believe our workflow provides an tractable step to fixing this. By focusing on model development more tightly tied to ecological expertise, we ague this workflow should broader the community that contributes to model development. As ecologists are increasingly expanding their computational toolkits, many field, and lab and other forms of `empirical' ecologists have the basic tools to follow this workflow to build models that better represent their ecological, and---most importantly---to interrogate them.

An universal workflow offers an opportunity to bridge statistical and process-based frameworks, integrating mechanistic knowledge and leveraging robust statistical approaches \citep[e.g.][]{rounce2020quantifying}. Process-based models would no longer be perceived as deterministic black boxes by other researchers but rather as robust statistical frameworks encapsulating both data structure and mechanistic knowledge. And it would also be an opportunity to spread the incorporation of mechanistic assumptions beyond the process-based modeling community.
In a world where machine learning is rapidly advancing, there is no point of sticking to traditional methods if no changes are made. Machine learning may surpass process-based models if the latter lack a robust estimation of their parameters and fall in a complexity trap, at the cost of their interpretability. Similarly, trend analysis, when the focus is on methodological controversies (due to the lack of an iterative workflow) rather than on a robust mechanistic foundation, offers no clear advantage over machine learning.

% emw03april: massively shorten?
Advancing ecology to where most researchers see models built more flexibly from their understanding of their ecological systems will not happen rapidly without a major shift in training, however. Much of ecology still divides the world into training for those focused on several groups. First, those who will conduct field and lab studies often learn a suite of particular tests matched to particular experiment designs or to simply match their predictor and response data to a flow diagram of variable types (add EXAMPLE and reference, something like: categorical $y$ and binary $y$ means ... I dunno, chi square?) and are left adrift later when asked to simulate any sort of simple model because they were never taught what a model fundamentally is. Instead, they are expected to collaborate with others when they need more complex models, which is a second group of ecological researchers---those who focus on complex statistical models and were thus trained more in model development, but often for highly specific applications (e.g., wildlife population estimates) where they may rarely build totally new generative models. Separate from these groups, process-based modelers do their own thing. 
% process-based modelers---often with a physical or ecophysiological background---mostly focus on simulating some mechanisms 
% hmm some people talk about simulation models (for process-based models), I agree but I'm afraid in would not be clear with the data simulation we try to highlight
%
Few of these groups have integrated data simulation into their statistical or scientific workflows, which is generally reserved as a form of training needed mainly by those specializing in theoretical ecology, who often solve analytical equations but rarely link to empirical data. While specialization is valuable, we argue the fundamental training in ecology has overly-siloed these groups and prevented more rapid progress.
